---
title: "scMaSigPro: Bridging maSigPro Analysis"
subtitle: "Guide to Applying maSigPro Methods with scMaSigPro"
author: "Priyansh Srivastava"
package: "scMaSigPro"
abstract: |
  `scMaSigPro` is an R package designed for the serial analysis of single-cell 
  RNA-seq (scRNA-Seq) data along inferred pseudotime. It builds upon the [`maSigPro`](https://www.bioconductor.org/packages/release/bioc/html/maSigPro.html)
  Bioconductor package to identify genes with significant expression changes across different 
  branching paths in a pseudotime-ordered scRNA-Seq dataset. This vignette illustrates
  the workflow of the original [`maSigPro`](https://www.bioconductor.org/packages/release/bioc/html/maSigPro.html) Bioconductor package of using `scMaSigPro`. 
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{scMaSigPro Basic Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(scMaSigPro)
```

## Introduction
`scMaSigPro` is a polynomial regression-based approach inspired by the [maSigPro](https://www.bioconductor.org/packages/release/bioc/html/maSigPro.html)
Bioconductor package tailored for scRNA-Seq data. It first discretizes single cell
expression along inferred pseudotime while preserving order. Afterwards it applies
the [maSigPro model](https://academic.oup.com/bioinformatics/article/22/9/1096/200371)
to pinpoint genes exhibiting significant expression profile differences among
branching paths and pseudotime.

## Overview of `maSigPro`
maSigPro (microarray significant profiles) is a bioconductor package. Originally
designed for microarray data analysis, maSigPro focuses on identifying genes
with significant expression profiles that vary among different experimental
groups, particularly in time-course experiments.

## Adaptation for Bulk-RNA Seq Data
In 2016, the maSigPro package was adapted for bulk-RNA seq data, a crucial
development given the distinct differences between microarray and RNA-Seq
technologies. Unlike the distributions typically seen in microarray data,
bulk-RNA Seq data generally follows a Negative Binomial Distribution, which is
more apt for count data like RNA-Seq, characterized by variance often exceeding
the mean. 

This adaptation brought in explicit theta parameterization, critical
for RNA-Seq data analysis as the theta parameter represents the dispersion in
the Negative Binomial Distribution. Such parameterization enhances the accuracy
in modeling gene expression variability, a key factor in identifying significant
changes in gene expression, thereby marking a significant step forward in the
application of maSigPro to the more complex RNA-Seq data.

## Set-p and Installation
### Installing `scMaSigPro`
Currently, `scMaSigPro` is available on GitHub and can be installed as follows:

```{r, "scMaSigPro-install", echo=TRUE, eval=FALSE}
# Use public PAT
publicPat <- "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"

# Install devtools if not already installed
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

# Install scMaSigPro
devtools::install_github("spriyansh/scMaSigPro",
  ref = "dev",
  auth_token = publicPat,
  build_vignettes = TRUE,
  build_manual = TRUE,
  upgrade = "never",
  force = TRUE,
  quiet = TRUE,
)
```

### Installing `maSigPro`
maSigPro can be easily installed from Bioconductor as follows:

```{r, "maSigPro-install", echo=TRUE, eval=FALSE}
# Install Bioconductor if not already installed
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# install maSigPro
BiocManager::install("maSigPro")
```

### Set up
```{r, "set up", echo=TRUE, eval=TRUE}
# load scMaSigPro
library(scMaSigPro)

# load maSigPro
library(maSigPro)

# We will also keep checking the output
library(testthat)
```

### Setting up the data
For this tutorial we will make use of the datasets from maSigPro and will try to
reproduce the same analysis pipeline with scMaSigPro. More details on the analysis
of the data are given in the original [`maSigPro` vignette-userguide](https://www.bioconductor.org/packages/release/bioc/vignettes/maSigPro/inst/doc/maSigProUsersGuide.pdf)

```{r, "get-data", echo=TRUE, eval=TRUE}
# load counts
data(data.abiotic)

# load Design file
data(edesign.abiotic)

# save with different names for scMaSigPro
count <- data.abiotic
metadata <- as.data.frame(edesign.abiotic)
```

Let's quickly view what each of the dataset has
```{r, "view-data", echo=TRUE, eval=TRUE}
print(data.abiotic[c(1:5),c(1:5)])

# load Design file
print(edesign.abiotic[c(1:5),c(1:5)])
```

Since we already have the groups binarized we will first convert them into a single colum
```{r, "non-binarize", echo=TRUE, eval=TRUE}
# Add group column
metadata$Group <- apply(metadata[,c(3:6)], 1, FUN = function(x){
    return(names(x[x == 1]))})

# Remove binary columns
metadata <- metadata[,!(colnames(metadata) %in% c("Control", "Cold", "Heat","Salt")), drop = FALSE]
```

## Creatin objects for maSigPro and scMaSigPro

```{r, "create-scmpObject", echo=TRUE, eval=TRUE}
scmp.ob <- create.scmp(
    counts = count,
    cell_data = metadata,
    pseudotime_colname = "Time",
    path_colname = "Group",
    use_as_bin = T
)
```

```{r}
expect_equal(eDense(scmp.ob), expected = data.abiotic)
```

## Creatin of design file 

```{r}
design <- make.design.matrix(edesign.abiotic, degree = 2)
```

```{r}
scmp.ob <- sc.set.poly(scmp.ob, poly_degree = 2)
```

```{r}
expect_equal(bAlloc(scmp.ob), expected = edesign.abiotic)
```

```{r}
expect_equal(scmp.ob@design@predictor, expected = design$dis)
```

## Pvector
```{r}
gc <- capture_output(fit <- p.vector(data.abiotic, design, Q = 0.05,
                                     MT.adjust = "BH", min.obs = 20))
```

```{r}
scmp.ob <- sc.p.vector(scmp.ob,
                       min.na = 20,
                       offset = F,
                       max_it = 25,
                       epsilon = 0.00001,
                       family = gaussian())
```


```{r}
expect_equal(matrix(scmp.ob@profile@p.vector, dimnames = list(names(scmp.ob@profile@p.vector), "p.value")), expected = as.matrix(fit$p.vector[,1, drop =FALSE]))
```

```{r}
pad <- scmp.ob@profile@p.adjusted
names(pad) <- NULL
expect_equal(pad, expected = fit$p.adjusted)
```

## t fit

```{r}
scmp.ob <- sc.t.fit(scmp.ob,
                       offset = F,
                       epsilon = 0.00001,
                       family = gaussian())
```

```{r}
gc <- capture.output(tstep <- T.fit(fit, step.method = "backward", alfa = 0.05))
```

```{r}
expect_equal(showSol(scmp.ob), expected = tstep$sol)
```

```{r}
expect_equal(showCoeff(scmp.ob), expected = tstep$coefficients)
```

```{r}
expect_equal(showGroupCoeff(scmp.ob), expected = tstep$group.coeffs)
```

```{r}
sigs <- get.siggenes(tstep, rsq = 0.6, vars = "groups")  
```

```{r}
scmp.ob <- sc.filter(scmp.ob, rsq = 0.6, vars = "groups")  
```

```{r}
expect_equal(scmp.ob@sig.genes@sig.genes$ColdvsControl,
             expected = sigs$summary$ColdvsControl)
```

```{r}
expect_equal(scmp.ob@sig.genes@sig.genes$ColdvsControl,
             expected = sigs$summary$ColdvsControl)
```

```{r}
STMDE66 <- data.abiotic[rownames(data.abiotic)=="STMDE66", ]
 PlotGroups (STMDE66, edesign = edesign.abiotic, show.fit = T,
             dis = design$dis, groups.vector = design$groups.vector)
```

```{r}
plotTrend(scmp.ob, "STMDE66", logs = F, pseudoCount = 0,
          smoothness = 1, significant = F)
```

```{r}
see.genes(sigs$sig.genes$ColdvsControl, show.fit = T, dis =design$dis,
cluster.method="hclust" ,cluster.data = 1, k = 9)
```
```{r}
scmp.ob <- sc.cluster.trend(scmp.ob, k = 9,
                            cluster.method = "hclust")
```

```{r}
plotTrendCluster(scmp.ob)
```

---

### Session Info
```{r, "Session Info"}
sessionInfo(package = "scMaSigPro")
```
